---
title: "Compensatory versus multiple-hurdle selection simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compensatory versus multiple-hurdle selection simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```


## The design question

A compensatory system applies a cutoff to a predictor composite: high scores on some predictors compensate for low scores on others, and applicants are ranked top-down on the composite. A multiple-hurdle system applies sequential cutoffs: an applicant must pass the cutoff on each predictor, or each stage, before reaching the next. Ock and Oswald (2018) formalise the trade-off between the two designs: compensatory systems often preserve more information and produce higher expected criterion performance, whereas multiple-hurdle systems can reduce administration costs because expensive stages are administered only to applicants who survive earlier stages. The cost-reliability trade-off is the substantive decision facing the analyst.

```{r}
library(personnelSelectionUtility)
```

## A four-predictor example

Following the type of simulation design used by Ock and Oswald (2018) and informed by the meta-analytic input matrices of Roth, Switzer, Van Iddekinge, and Oh (2011), suppose the predictors are cognitive ability, structured interview, conscientiousness, and biodata. The criterion is job performance.

```{r}
Rxx <- matrix(c(
  1.00, .31, .03, .37,
  .31, 1.00, .13, .16,
  .03, .13, 1.00, .51,
  .37, .16, .51, 1.00
), 4, 4, byrow = TRUE)
validities <- c(.37, .35, .16, .23)
```

The intercorrelations and validities are illustrative rather than reproductions of any single published meta-analysis. The qualitative pattern, however, mirrors what most contemporary meta-analyses report: cognitive ability has the highest validity but moderate correlation with the structured interview, conscientiousness is largely independent of cognitive ability, and biodata correlates substantially with conscientiousness while contributing additional incremental validity.

## Compensatory selection

In the compensatory system, all applicants complete all predictors and are selected top-down on a composite. The expected criterion gain follows the Naylor-Shine (1965) logic, with the validity coefficient replaced by the composite validity computed from the predictor intercorrelations and the weighting scheme. With predictor correlation matrix $\mathbf{R}_{XX}$, predictor-criterion validities $\mathbf{r}_{XY}$, and weight vector $\mathbf{w}$, the composite-criterion correlation follows the Lord and Novick (1968) formula

$$
r_{C,Y} \;=\; \frac{\mathbf{w}^{\top} \mathbf{r}_{XY}}{\sqrt{\mathbf{w}^{\top} \mathbf{R}_{XX}\, \mathbf{w}}}.
$$

```{r}
comp <- compensatory_selection(
  predictor_cor      = Rxx,
  validities         = validities,
  weights            = rep(1, 4),
  selection_ratio    = .20,
  n_applicants       = 500,
  cost_per_applicant = 1000,
  sdy                = 60000
)
comp
```

Equal weighting is rarely optimal, but it is a reasonable default when validity differences are small or uncertain (Bobko, Roth, & Buster, 2007; Wainer, 1976). When validities are well established, weighting by validity (or by the optimal regression weights derived from the full predictor-criterion correlation matrix) yields a higher composite validity at the cost of greater sample-to-sample weight instability.

## Staged multiple-hurdle selection

Now suppose the first stage is a cheaper composite of cognitive ability, conscientiousness, and biodata. The second stage is a structured interview administered only to applicants who pass the first stage. The first stage retains $25\%$ of applicants; the interview retains $80\%$ of those, giving an expected joint selection ratio near $.20$. This staged logic was examined formally by Sackett and Roth (1996) and is a natural representation of the operational reality in many high-volume selection contexts.

```{r}
R <- rbind(cbind(Rxx, validities), c(validities, 1))

hurdle <- multiple_hurdle_selection_staged(
  stage_predictors       = list(c(1, 3, 4), 2),
  stage_selection_ratios = c(.25, .80),
  R                      = R,
  n_sim                  = 5000,
  seed                   = 123,
  n_applicants           = 500,
  cost_per_stage         = c(100, 900),
  sdy                    = 60000
)
hurdle
```

The Monte Carlo implementation generates simulated applicants from the multivariate normal distribution defined by `R` and applies the staged cutoffs; the realised joint selection ratio and expected criterion performance are estimated from the simulation. Larger `n_sim` reduces simulation error at the cost of computation time. The vignette uses small values for illustrative speed; in production analyses values of $50,000$ or higher are recommended.

## Direct comparison

The convenience wrapper `compare_selection_systems_staged()` performs both calculations with the same inputs and returns a comparison object.

```{r}
comparison <- compare_selection_systems_staged(
  predictor_cor                  = Rxx,
  validities                     = validities,
  compensatory_weights           = rep(1, 4),
  compensatory_selection_ratio   = .20,
  stage_predictors               = list(c(1, 3, 4), 2),
  stage_selection_ratios         = c(.25, .80),
  n_sim                          = 5000,
  seed                           = 123,
  n_applicants                   = 500,
  compensatory_cost_per_applicant = 1000,
  hurdle_cost_per_stage          = c(100, 900),
  sdy                            = 60000
)
comparison
```

The comparison object reports differences in expected criterion performance, joint selection ratio, and net utility.

```{r}
c(
  expected_z_difference  = comparison$expected_criterion_z_difference,
  net_utility_difference = comparison$net_utility_difference
)
```

## Cost-reliability trade-off

The substantive decision is not which system has higher expected performance but whether the performance advantage of the more complete compensatory system offsets its additional cost. Ock and Oswald (2018) recommend exploring conditions rather than relying on a single point estimate. One way to study this is to vary $SD_y$ (which determines the dollar value of each unit of expected gain) and the relative cost of the hurdle system.

```{r}
sdy_values <- c(20000, 40000, 60000)
hurdle_stage2_cost <- c(200, 500, 900)

out <- expand.grid(sdy = sdy_values, interview_cost = hurdle_stage2_cost)
out$net_utility_difference <- NA_real_

for (i in seq_len(nrow(out))) {
  cmp <- compare_selection_systems_staged(
    predictor_cor                  = Rxx,
    validities                     = validities,
    compensatory_selection_ratio   = .20,
    stage_predictors               = list(c(1, 3, 4), 2),
    stage_selection_ratios         = c(.25, .80),
    n_sim                          = 3000,
    seed                           = 100 + i,
    n_applicants                   = 500,
    compensatory_cost_per_applicant = 1000,
    hurdle_cost_per_stage          = c(100, out$interview_cost[i]),
    sdy                            = out$sdy[i]
  )
  out$net_utility_difference[i] <- cmp$net_utility_difference
}

out
```

Positive values indicate that the compensatory system has higher net utility. Negative values indicate that the hurdle system is favoured under that cost and $SD_y$ scenario. The pattern is the one Ock and Oswald (2018) emphasise: at low $SD_y$ and high stage-2 cost, the hurdle system can dominate on net utility despite producing lower expected per-hire performance.

## Offer rejection: Murphy (1986)

A correction frequently omitted from utility analyses is the effect of rejected job offers. Murphy (1986), building on Hogarth and Einhorn (1976), showed that when top candidates reject offers, the utility of selection tests is overstated because the actually hired group has a lower expected predictor score than the offered group. The function `offer_rejection_adjustment()` implements three modes consistent with Murphy's analysis: uniform random rejection, rejection correlated with predictor score, and selectively higher rejection at the top of the distribution.

```{r}
# First compute the expected standardised score among offered candidates:
z_offered <- selected_mean_z(.20)

# Adverse selection (correlated mode): top candidates are more likely to decline,
# captured by a negative correlation between standardised quality and acceptance.
offer_rejection_adjustment(
  expected_z_offered     = z_offered,
  mode                   = "correlated",
  acceptance_rate        = .70,
  rho_quality_acceptance = -0.20,
  n_offered              = 100
)
```

The substantive case for the *correlated* mode is well established empirically: candidates with stronger profiles tend to have more outside options, so the probability of accepting a given offer correlates negatively with the predictor score. Sturman (2001) used $\rho_{\text{quality, acceptance}} = -0.20$ and an acceptance rate of $.70$ in his comprehensive model. Under these conditions, the realised mean predictor score among the *hired* group is materially lower than the inverse-Mills mean among the *offered* group, and the utility estimate must be adjusted downward.

## Adverse impact and the validity-diversity dilemma

Selection systems that maximise composite validity may produce subgroup hire-rate differences that violate the four-fifths threshold articulated in the Uniform Guidelines on Employee Selection Procedures (1978). Pyburn, Ployhart, and Kravitz (2008) framed this as the *validity-diversity dilemma*: the predictors with the highest validity for job performance also tend to produce the largest mean differences between demographic subgroups. The function `adverse_impact_ratio()` computes the four-fifths comparison from group-specific selection ratios.

```{r}
# adverse_impact_ratio() takes individual-level selection outcomes and group labels;
# it computes the selection rate per group and the four-fifths ratio relative to
# the group with the highest rate.
selected <- c(1, 1, 0, 1, 0, 1, 0, 1, 0,
              1, 0, 0, 1, 0, 0, 0, 1, 0)
group    <- c(rep("Reference", 9), rep("Focal", 9))
adverse_impact_ratio(selected, group)
```

De Corte, Lievens, and Sackett (2007) developed the Pareto-optimal solution to this trade-off: rather than choosing one weighting scheme, the analyst characterises the entire frontier of weighting schemes that are Pareto-optimal in the validity-diversity plane, leaving the final choice to organisational stakeholders. The package implements this through `pareto_frontier()` and the related `utility_fairness_frontier()`.

```{r}
# pareto_frontier() is a general Pareto-membership indicator: given a matrix of
# objectives (rows = alternatives, columns = objectives to maximise), it returns
# a logical vector flagging the non-dominated alternatives. The validity-diversity
# trade-off in selection systems is one application; below we evaluate six candidate
# weighting schemes on composite validity and four-fifths fairness.
candidates <- data.frame(
  scheme   = c("CA only", "CA + interview",
               "Equal weights", "Validity weights",
               "Pareto-optimal #1", "Pareto-optimal #2"),
  validity = c(.51, .55, .50, .56, .53, .54),
  fairness = c(.62, .68, .73, .65, .76, .80)
)
candidates$pareto <- pareto_frontier(
  objectives = candidates[, c("validity", "fairness")],
  maximize   = TRUE
)
candidates
```

The robustness and shrinkage properties of Pareto-optimal solutions across cross-validation samples were studied by Song, Wee, and Newman (2017) and De Corte, Sackett, and Lievens (2020); the analyst should report both the in-sample frontier and an out-of-sample shrinkage estimate when sample sizes are modest.

## Multi-attribute utility

When the criterion is genuinely multi-attribute (task performance, contextual performance, and counterproductive work behaviour, for example, following the taxonomies of Borman and Motowidlo (1993) and Rotundo and Sackett (2002)) the appropriate framework is multi-attribute utility analysis (Roth, 1994; Roth & Bobko, 1997), which decomposes overall utility as a weighted sum of attribute-specific utilities under the assumption of mutual preferential independence (Keeney & Raiffa, 1976).

```{r}
# Two candidate selection systems evaluated on three attributes (task,
# contextual, CWB avoidance), with values on a common 0-100 scale:
values <- matrix(c(
  80, 60, 90,
  70, 75, 70
), nrow = 2, byrow = TRUE,
   dimnames = list(c("System A", "System B"),
                   c("task", "contextual", "cwb_avoidance")))

multiattribute_utility(
  values  = values,
  weights = c(.50, .30, .20)
)
```

The substantive case for multi-attribute utility analysis, formalised by Roth and Bobko (1997), is that aggregating dollar values of heterogeneous outcomes into a single $SD_y$ disguises the underlying preference structure of the organisation. Reporting attribute-specific utilities preserves the information needed to negotiate trade-offs explicitly.

## Risk-adjusted utility

Bhattacharya and Wright (2005) introduced risk-adjustment to selection utility, treating future utility flows as stochastic and pricing the risk through a real-options or certainty-equivalent framework. The function `risk_adjusted_utility()` adjusts the expected utility by a risk premium that reflects the variance of the per-period utility flows over the planning horizon.

```{r}
# The mean-variance risk-adjusted score subtracts a penalty proportional to the
# variance of utility. Because monetary utilities are often in the millions, the
# risk_aversion parameter is typically very small (e.g., 1e-6 to 1e-5). The
# example below uses the compensatory net utility computed earlier.
risk_adjusted_utility(
  expected_utility = comparison$compensatory$net_utility,
  utility_sd       = abs(comparison$compensatory$net_utility) * .30,
  risk_aversion    = 1e-6
)
```

Risk adjustment is most consequential for long planning horizons, uncertain validity coefficients, and volatile labour markets. For analyses spanning a single year, the risk premium is typically small relative to point-estimate uncertainty in $SD_y$ and validity, and reporting a Monte Carlo posterior distribution through `utility_monte_carlo()` will usually convey the same information more transparently.

## How to proceed in applied work

1. Define the actual stages used by the organisation; do not collapse a real multiple-hurdle system into a compensatory model for computational convenience.
2. Estimate costs separately by stage. A single average cost per applicant misrepresents the cost-reliability trade-off when early stages are inexpensive screens and late stages are expensive assessments.
3. Use realistic validity coefficients corrected for measurement error and range restriction *before* utility analysis.
4. Compare expected criterion performance and monetary utility; the two can diverge under cost differentials, and Ock and Oswald (2018) document that hurdle systems can dominate on net utility while losing on per-hire performance.
5. Vary $SD_y$, costs, and selection ratios. A result that changes sign under plausible inputs should be reported as fragile; sensitivity analysis is mandatory rather than optional.
6. Use larger `n_sim` values for final analyses; the vignette examples use small values for execution speed.
7. Adjust for offer rejection using `offer_rejection_adjustment()` whenever competitive labour markets make the assumption of universal acceptance untenable.
8. Report the validity-diversity Pareto frontier, not a single weighting, when the decision has legal or fairness implications. The frontier preserves the trade-off information needed for organisational negotiation.
9. Use `multiattribute_utility()` when the criterion is multi-attribute and the organisation has explicit weights, rather than aggregating into a single $SD_y$.
10. Report risk-adjusted utility or, equivalently, a Monte Carlo posterior distribution of $\Delta U$ for long planning horizons or volatile inputs.

## References

Bhattacharya, M., & Wright, P. M. (2005). Managing human assets in an uncertain world: Applying real options theory to HRM. *International Journal of Human Resource Management*, *16*, 929--948.

Bobko, P., Roth, P. L., & Buster, M. A. (2007). The usefulness of unit weights in creating composite scores: A literature review, application to content validity, and meta-analysis. *Organizational Research Methods*, *10*, 689--709.

Borman, W. C., & Motowidlo, S. J. (1993). Expanding the criterion domain to include elements of contextual performance. In N. Schmitt, W. C. Borman, & Associates (Eds.), *Personnel selection in organizations* (pp. 71--98). Jossey-Bass.

De Corte, W., Lievens, F., & Sackett, P. R. (2007). Combining predictors to achieve optimal trade-offs between selection quality and adverse impact. *Journal of Applied Psychology*, *92*, 1380--1393.

De Corte, W., Sackett, P. R., & Lievens, F. (2020). Robustness, sensitivity, and sampling variability of Pareto-optimal selection system solutions to address the quality-diversity trade-off. *Organizational Research Methods*, *23*, 511--535.

Hogarth, R. M., & Einhorn, H. J. (1976). Optimal strategies for personnel selection when candidates can reject job offers. *Journal of Business*, *49*, 479--495.

Keeney, R. L., & Raiffa, H. (1976). *Decisions with multiple objectives: Preferences and value tradeoffs*. Wiley.

Murphy, K. R. (1986). When your top choice turns you down: Effect of rejected offers on the utility of selection tests. *Psychological Bulletin*, *99*, 133--138.

Naylor, J. C., & Shine, L. C. (1965). A table for determining the increase in mean criterion score obtained by using a selection device. *Journal of Industrial Psychology*, *3*, 33--42.

Ock, J., & Oswald, F. L. (2018). The utility of personnel selection decisions: Comparing compensatory and multiple-hurdle selection models. *Journal of Personnel Psychology*, *17*(4), 172--182.

Pyburn, K. M., Ployhart, R. E., & Kravitz, D. A. (2008). The diversity-validity dilemma: Overview and legal context. *Personnel Psychology*, *61*, 143--151.

Roth, P. L. (1994). Multi-attribute utility analysis using the PROMES approach. *Journal of Business and Psychology*, *9*, 69--80.

Roth, P. L., & Bobko, P. (1997). A research agenda for multi-attribute utility analysis in human resource management. *Human Resource Management Review*, *7*, 341--368.

Roth, P. L., Switzer, F. S., Van Iddekinge, C. H., & Oh, I. S. (2011). Toward better meta-analytic matrices: How input values can affect research conclusions in human resource management simulations. *Personnel Psychology*, *64*, 899--935.

Rotundo, M., & Sackett, P. R. (2002). The relative importance of task, citizenship, and counterproductive performance to global ratings of job performance: A policy-capturing approach. *Journal of Applied Psychology*, *87*, 66--80.

Sackett, P. R., & Roth, L. (1996). Multi-stage selection strategies: A Monte Carlo investigation of effects on performance and minority hiring. *Personnel Psychology*, *49*, 549--572.

Song, Q. C., Wee, S., & Newman, D. A. (2017). Diversity shrinkage: Cross-validating Pareto-optimal weights to enhance diversity via hiring practices. *Journal of Applied Psychology*, *102*, 1636--1657.

Sturman, M. C. (2001). Utility analysis for multiple selection devices and multiple outcomes. *Journal of Human Resource Costing and Accounting*, *6*(2), 9--28.

Wainer, H. (1976). Estimating coefficients in linear models: It don't make no nevermind. *Psychological Bulletin*, *83*, 213--217.
