---
title: "Continuous and monetary utility: Naylor-Shine, BCG, Boudreau, and Sturman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Continuous and monetary utility: Naylor-Shine, BCG, Boudreau, and Sturman}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```


## From success ratios to continuous performance

Taylor-Russell utility dichotomizes the criterion. Naylor and Shine (1965) keep the criterion continuous and compute expected criterion gain in standard deviation units. This is usually a better match when performance is a continuous construct.

```{r}
library(personnelSelectionUtility)
naylor_shine(validity = .35, selection_ratio = .20)
```

The expected gain has two steps:

1. compute the expected standardized predictor score among selected applicants;
2. multiply it by the validity coefficient.

```{r}
selected_mean_z(.20)
.35 * selected_mean_z(.20)
```

## Brogden-Cronbach-Gleser monetary utility

Brogden (1946, 1949) and Cronbach and Gleser (1965) convert the standardized criterion gain into utility by multiplying by `SD_y`, the standard deviation of job performance value. Schmidt, Hunter, McKenzie, and Muldrow (1979) popularized the organizational implications of this formulation for personnel selection.

```{r}
bcg_utility(
  validity = .35,
  selection_ratio = .20,
  sdy = 50000,
  n_selected = 100,
  tenure = 3,
  cost = 75000
)
```

## The baseline problem

The classical expression compares against random selection. Sturman (2001) argues that the realistic baseline is often another selection system, not random choice. The package represents this with `baseline_validity` and `baseline_selection_ratio`.

```{r}
random_baseline <- bcg_utility(
  validity = .35,
  selection_ratio = .20,
  sdy = 50000,
  n_selected = 100,
  tenure = 3,
  cost = 75000
)

operating_baseline <- bcg_utility(
  validity = .35,
  baseline_validity = .20,
  selection_ratio = .20,
  sdy = 50000,
  n_selected = 100,
  tenure = 3,
  cost = 75000
)

c(random = random_baseline$net_utility,
  operating = operating_baseline$net_utility)
```

## Estimating `SD_y`

Holling (1998) shows why `SD_y` is a central vulnerability in utility analysis. Small changes in `SD_y`, especially when objective performance distributions are skewed or outliers are present, can change monetary utility dramatically. The package includes several helpers.

Percentile judgments use the familiar 15th-to-85th percentile approximation.

```{r}
sdy_percentile(p15 = 60000, p85 = 140000)
```

Proportional rules express `SD_y` as a fraction of pay or output value.

```{r}
sdy_proportional(mean_pay = 80000, multiplier = .40)
sdy_proportional(mean_pay = 80000, multiplier = .70)
```

Observed monetary criterion data can be used directly.

```{r}
sdy_observed(c(90000, 110000, 85000, 150000, 125000))
```

## Boudreau-style extensions

Boudreau (1983, 1991) extends the basic model by adding discounting, taxes, value multipliers, and employee flows. In `personnelSelectionUtility`, the preferred argument names are `n_by_period` and `cost_by_period`.

```{r}
boudreau_utility(
  validity = .35,
  baseline_validity = .20,
  selection_ratio = .20,
  sdy = 50000,
  n_by_period = c(100, 90, 80, 70),
  contribution_margin = .30,
  tax_rate = .25,
  discount_rate = .08,
  cost_by_period = c(75000, 10000, 10000, 10000)
)
```

If the expected incremental standardized gain comes from an external model, pass it directly through `delta_z_y`.

```{r}
boudreau_utility(
  delta_z_y = .25,
  sdy = 50000,
  n_by_period = c(100, 90, 80),
  discount_rate = .08,
  cost_by_period = c(75000, 10000, 10000)
)
```

## Multiple predictors and multiple outcomes

Sturman (2001) generalizes utility analysis to multiple selection devices and multiple outcomes. The key move is to evaluate the incremental validity of a new predictor system over the current system, while allowing the criterion to be a weighted composite of several outcomes.

```{r}
S11 <- matrix(c(1, .30, .30, 1), 2, 2)
S12 <- matrix(c(.30, .20, .25, .15), 2, 2)
S22 <- matrix(c(1, .40, .40, 1), 2, 2)

restricted_canonical_validity(S11, S12, S22, criterion_weights = c(.6, .4))
```

To evaluate whether adding a predictor is worthwhile, use `incremental_validity()` and then feed the resulting increment into a utility model.

```{r}
Rxx <- matrix(c(1, .30, .20,
                .30, 1, .25,
                .20, .25, 1), 3, 3, byrow = TRUE)
Rxy <- matrix(c(.30, .20,
                .25, .15,
                .10, .35), 3, 2, byrow = TRUE)
Ryy <- matrix(c(1, .40, .40, 1), 2, 2)

incremental_validity(
  predictor_cor = Rxx,
  predictor_criterion_cor = Rxy,
  criterion_cor = Ryy,
  criterion_weights = c(.6, .4),
  baseline_predictors = 1:2,
  added_predictors = 3
)
```

## How to proceed in applied work

1. Start with `naylor_shine()` if you only need expected criterion gain.
2. Use `bcg_utility()` for transparent monetary utility, but do not stop there.
3. Replace random selection with the actual operating baseline whenever possible.
4. Estimate `SD_y` using more than one method and report a sensitivity range.
5. Use `boudreau_utility()` when utility spans several periods or when costs/returns occur at different times.
6. Use `restricted_canonical_validity()` and `incremental_validity()` when the current selection system already includes predictors.
7. Report the input assumptions; utility estimates are only as credible as the validity, `SD_y`, cost, and tenure assumptions.

## References

Boudreau, J. W. (1983). Economic considerations in estimating the utility of human resource productivity improvement programs. *Personnel Psychology*, 36, 551--576.

Boudreau, J. W. (1991). Utility analysis for decisions in human resource management. In M. D. Dunnette & L. M. Hough (Eds.), *Handbook of industrial and organizational psychology* (Vol. 2, pp. 621--745). Consulting Psychologists Press.

Brogden, H. E. (1946). On the interpretation of the correlation coefficient as a measure of predictive efficiency. *Journal of Educational Psychology*, 37, 65--76.

Brogden, H. E. (1949). When testing pays off. *Personnel Psychology*, 2, 171--183.

Cronbach, L. J., & Gleser, G. C. (1965). *Psychological tests and personnel decisions* (2nd ed.). University of Illinois Press.

Holling, H. (1998). Utility analysis of personnel selection: An overview and empirical study based on objective performance measures. *Methods of Psychological Research Online*, 3(1), 5--24.

Naylor, J. C., & Shine, L. C. (1965). A table for determining the increase in mean criterion score obtained by using a selection device. *Journal of Industrial Psychology*, 3, 33--42.

Schmidt, F. L., Hunter, J. E., McKenzie, R. C., & Muldrow, T. W. (1979). Impact of valid selection procedures on work-force productivity. *Journal of Applied Psychology*, 64, 609--626.

Sturman, M. C. (2001). Utility analysis for multiple selection devices and multiple outcomes. *Journal of Human Resource Costing and Accounting*, 6(2), 9--28.
