---
title: "Utility-analysis taxonomy for personnel selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Utility-analysis taxonomy for personnel selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```


## Purpose

The package is organized around a diagnostic question: **what kind of selection problem are you analyzing?** Many disagreements in utility analysis arise because a method designed for one type of problem is used for another. Taylor and Russell (1939) framed utility as a classificatory success-ratio problem. Naylor and Shine (1965) moved the focus to expected criterion gain in standard-deviation units. Brogden (1946, 1949) and Cronbach and Gleser (1965) developed the decision-theoretic monetary formulation. Boudreau (1983, 1991), Holling (1998), Sturman (2001), Thomas, Owen, and Gunst (1977), and Ock and Oswald (2018) each added important corrections or extensions.

The taxonomy used here crosses two dimensions:

1. **Criterion scale**: is the criterion treated as continuous/monetary, or dichotomized into success/failure?
2. **Selection structure**: is selection compensatory, based on a composite score, or conjunctive/multiple-hurdle, based on passing multiple cutoffs or stages?

```{r}
library(personnelSelectionUtility)
model_taxonomy()
```

This yields four practical cells.

| Criterion scale | Compensatory selection | Multiple-hurdle selection |
|---|---|---|
| Dichotomized / classificatory | Taylor-Russell on one predictor or a composite | Thomas-Owen-Gunst multivariate Taylor-Russell |
| Continuous / monetary | Naylor-Shine, BCG, Boudreau, Sturman incremental validity | Simulation or stage-wise utility calculations |

## Why this matters

A selection system can have the same predictors but a different utility model depending on the decision rule. If cognitive ability, conscientiousness, biodata, and interview ratings are added into one composite and applicants are selected top-down, the model is compensatory. If applicants must first pass a cheap screening composite and only then pass an interview stage, the model is staged multiple-hurdle.

The distinction is not cosmetic. Ock and Oswald (2018) emphasize the cost-reliability trade-off: a compensatory composite often uses more information and yields higher expected criterion performance, whereas a multiple-hurdle system can be cheaper because expensive stages are administered only to applicants who survive earlier screens. Sturman (2001) adds a second major warning: classical utility analyses often compare a new procedure to random selection, even though the realistic baseline is usually an existing selection system.

## Argument naming

The package uses readable R argument names while keeping close correspondence with the notation used in the literature.

```{r}
head(argument_glossary(), 12)
```

The most important names are:

`base_rate`: population proportion successful before selection; Taylor-Russell notation often uses `BR`.

`selection_ratio`: proportion selected by a single cutoff or composite; classical notation uses `SR`.

`selection_ratios`: vector of marginal selection ratios in multiple-predictor or multiple-stage models.

`joint_selection_ratio`: overall conjunctive selection ratio after all cutoffs.

`validity`: predictor-criterion validity coefficient, usually `r_xy`.

`validities`: vector of predictor-criterion correlations.

`sdy`: standard deviation of job performance in monetary or criterion units, usually `SD_y`.

`n_applicants`: number of applicants assessed by a system.

`n_selected`: number selected or treated.

## Recommended workflow

### Step 1: Specify the selection decision

Start by writing down the rule used in practice. Is selection based on a single test, a composite, several simultaneous cutoffs, or a staged process? Do not choose a utility formula before specifying the rule.

### Step 2: Specify the criterion scale

If the criterion is success/failure, use Taylor-Russell style functions. If the criterion is continuous or monetary, use Naylor-Shine, BCG, Boudreau, or simulation functions.

### Step 3: Specify the baseline

Following Sturman (2001), avoid treating random selection as the default comparator unless it is genuinely the decision alternative. Use `baseline_validity` in `bcg_utility()` or `boudreau_utility()` when the organization already has an operating procedure.

### Step 4: Estimate or triangulate `SD_y`

Holling (1998) highlights that `SD_y` is often the weakest link in monetary utility analysis. Use multiple estimates when possible: observed criterion data, percentile judgments, proportional rules, or job-analysis-based estimates.

### Step 5: Report uncertainty and sensitivity

Ock and Oswald (2018) show that utility estimates can vary substantially across samples. Use `utility_monte_carlo()`, `sensitivity_grid()`, and `break_even_validity()` to avoid reporting a single deterministic point estimate.

## Minimal examples by model family

### Dichotomous success criterion

For a single predictor or a composite summarized by one validity coefficient, use `tr_classic()`.

```{r}
tr_classic(base_rate = .50, selection_ratio = .20, validity = .35)
```

For multiple simultaneous cutoffs, use `tr_multivariate()`.

```{r}
R <- matrix(c(
  1.00, .30, .40,
  .30, 1.00, .35,
  .40, .35, 1.00
), nrow = 3, byrow = TRUE)
tr_multivariate(selection_ratios = c(.50, .50), base_rate = .50, R = R)
```


### AUC and effect-size conversions

In contemporary selection research, especially when using algorithmic or
classification models, predictive performance may be reported as AUC rather
than as a validity coefficient. The package therefore separates three
conceptually distinct conversions. First, `auc_to_rank_biserial()` returns the
dominance or rank-biserial summary `2 * AUC - 1`, which follows from the
interpretation of AUC as the probability of a favorable ordering of one positive
and one negative case (Hanley & McNeil, 1982; Kerby, 2014). Second,
`auc_to_d_equal_variance()` converts AUC to Cohen's `d` under the
equal-variance binormal model (Rice & Harris, 2005; Salgado, 2018). Third,
`auc_to_point_biserial()` converts that `d` to a point-biserial correlation for
a specified `base_rate`, making the base-rate dependence explicit.

```{r}
auc_to_rank_biserial(.75)
auc_to_d_equal_variance(.75)
auc_to_point_biserial(.75, base_rate = c(.50, .30, .20, .10))
```

Use these conversions as bridges between reported classification performance
and utility-analysis inputs, not as assumption-free substitutes for validation
studies. If the selection criterion is binary and the available evidence is AUC,
report which conversion was used and whether a base rate was assumed.

### Continuous or monetary criterion

For expected standardized criterion gain without money, use `naylor_shine()`.

```{r}
naylor_shine(validity = .35, selection_ratio = .20)
```

For monetary utility, use `bcg_utility()` as a transparent baseline model.

```{r}
bcg_utility(
  validity = .35,
  selection_ratio = .20,
  sdy = 50000,
  n_selected = 100,
  tenure = 3,
  cost = 75000
)
```

If you need discounting, costs by period, taxes, contribution margin, or employee flows, use `boudreau_utility()`.

```{r}
boudreau_utility(
  validity = .35,
  baseline_validity = .20,
  selection_ratio = .20,
  sdy = 50000,
  n_by_period = c(100, 90, 80),
  contribution_margin = .30,
  tax_rate = .25,
  discount_rate = .08,
  cost_by_period = c(75000, 10000, 10000)
)
```

## Reporting checklist

A complete utility analysis should report: the selection rule, the criterion scale, the base rate if a classificatory model is used, the selection ratio, all validity coefficients and their sources, how `SD_y` was estimated, whether validity and `SD_y` were corrected or not, the baseline comparator, costs, time horizon, discounting, uncertainty intervals, and sensitivity analyses.

## References

Boudreau, J. W. (1983). Economic considerations in estimating the utility of human resource productivity improvement programs. *Personnel Psychology*, 36, 551--576.

Boudreau, J. W. (1991). Utility analysis for decisions in human resource management. In M. D. Dunnette & L. M. Hough (Eds.), *Handbook of industrial and organizational psychology* (Vol. 2, pp. 621--745). Consulting Psychologists Press.

Brogden, H. E. (1946). On the interpretation of the correlation coefficient as a measure of predictive efficiency. *Journal of Educational Psychology*, 37, 65--76.

Brogden, H. E. (1949). When testing pays off. *Personnel Psychology*, 2, 171--183.

Cronbach, L. J., & Gleser, G. C. (1965). *Psychological tests and personnel decisions* (2nd ed.). University of Illinois Press.


Hanley, J. A., & McNeil, B. J. (1982). The meaning and use of the area under a receiver operating characteristic (ROC) curve. *Radiology*, 143(1), 29--36.

Kerby, D. S. (2014). The simple difference formula: An approach to teaching nonparametric correlation. *Comprehensive Psychology*, 3, 11.IT.3.1.

Rice, M. E., & Harris, G. T. (2005). Comparing effect sizes in follow-up studies: ROC area, Cohen's d, and r. *Law and Human Behavior*, 29(5), 615--620.

Salgado, J. F. (2018). Transforming the area under the normal curve (AUC) into Cohen's d, Pearson's r_pb, odds-ratio, and natural log odds-ratio: Two conversion tables. *The European Journal of Psychology Applied to Legal Context*, 10(1), 35--47.

Holling, H. (1998). Utility analysis of personnel selection: An overview and empirical study based on objective performance measures. *Methods of Psychological Research Online*, 3(1), 5--24.

Naylor, J. C., & Shine, L. C. (1965). A table for determining the increase in mean criterion score obtained by using a selection device. *Journal of Industrial Psychology*, 3, 33--42.

Ock, J., & Oswald, F. L. (2018). The utility of personnel selection decisions: Comparing compensatory and multiple-hurdle selection models. *Journal of Personnel Psychology*, 17(4), 172--182.

Sturman, M. C. (2001). Utility analysis for multiple selection devices and multiple outcomes. *Journal of Human Resource Costing and Accounting*, 6(2), 9--28.

Taylor, H. C., & Russell, J. T. (1939). The relationship of validity coefficients to the practical effectiveness of tests in selection. *Journal of Applied Psychology*, 23, 565--578.

Thomas, J. G., Owen, D. B., & Gunst, R. F. (1977). Improving the use of educational tests as selection tools. *Journal of Educational Statistics*, 2(1), 55--77.
